Lens Modeling in \sw

### 2014-06-15 text/partA.tex
editing

The first discoveries of gravitational lenses, in each case a galaxy
lensing a background quasar into multiple images
\citep{1979Natur.279..381W,1980Natur.285..641W}, brought in their wake
the first work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y}.  For one of the
lenses, mass modelling predicted that one of the lensed quasar images
seen would split further into two at higher resolution, which was
crucial in subsequently confirming that the object is indeed an
example of lensing.  It had long been expected that galaxies must
sometimes cause lensed images \citep{1937ApJ....86..217Z}, and it even
been argued that the phenomenon could help measure cosmological
parameters \citep{1964MNRAS.128..307R,1966MNRAS.132..101R}.  But
apparently nobody was expecting that lenses would need detailed
modelling.  In hindsight, the reason is not difficult to understand.
The scale of image separations is set by the angular Einstein radius
\begin{equation}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2}
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  For
stellar-mass lenses, the image separation is much larger than the size
of the lensing mass, so lenses are well approximated by point masses.
In galaxy lenses, the image separation is comparable to the size of
the galaxy; typically the lensed images are seen through the galaxy
halo.  Consequently, the lensed images depend on the detailed mass
distribution of the lensing galaxy.  Once observations of the lensed
images became available, the need to model the lensing-mass
distribution in detail became obvious.

Nowadays, detailed modelling of the mass distribution is part of any
research using multiply-imaging lenses.  Many works combine multiple
lenses with different mass ranges and other properties.  The largest
single study so far used 58 lenses \citep{2009ApJ...703L..51K} to
uncover the distribution of dark matter around galaxies.
\cite{2011ApJ...740...97L} combined lens models of 21 galaxies with
models of their stellar populations, to study the relation between
stars and dark matter in galaxies.  \cite{2014MNRAS.437..600S}
modelled 18 time-delay lenses together to infer cosmological
parameters.  But lens modelling has remained a separate process from
lens discovery.

The question now arises: can lenses be found and modeled
automatically in large surveys?  Work using software robots
\citep{2009ApJ...694..924M} gives some interesting and unexpected
results.  In clean lensing system in uncrowded fields with high
signal-to-noise, the robots do very well.  In most situations,
however, robots miss lenses (low completeness) or contaminate the
results with non-lenses (low purity).  Robots can be made to
prioritize completeness or purity, but they cannot deliver both.

In response to the lessons learned from automated lens searches, the
{\em Spacewarps\/} project\footnote{\tt http://www.spacewarps.org}
 has been launched.  {\em Spacewarps\/} is
part of the {\em Zooniverse\/} family of citizen-science
projects,\footnote{\tt http://www.zooniverse.org} where members of the
public are invited to analyze different kinds of scientific data which
are too difficult for robots and too large for specialists.  In {\em
 Spacewarps\/} itself, survey sky images are presented for volunteers to identify lens candidates.  The first tranche of survey data, introduced from May 2013, is an area of $\sim172$ square
degrees (or a quarter of a percent of the sky) from the Canada-France-Hawaii Telescope Legacy Survey.
 The survey is divided into patches of $440\times 440$ pixels, each of
which is seen by ten volunteers.  Simulated lenses are mixed in with
the data, both to help train volunteers on what to look for, and to
estimate completeness and purity.
These data had previously been analyzed by robots \cite{Gavazzi2012, More2012ApJ}.  
  A list of candidates is being
processed, but intermediate results show some very good candidates, as
well as re-discoveries of candidates previously found robotically.

>>>>

The first discoveries of gravitational lenses, in each case a galaxy
lensing a background quasar into multiple images
\citep{1979Natur.279..381W,1980Natur.285..641W}, brought in their wake
the first work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y}.  For one of the
lenses, mass modelling predicted that one of the lensed quasar images
seen would split further into two at higher resolution, which was
crucial in subsequently confirming that the object is indeed an
example of lensing.  It had long been expected that galaxies must
sometimes cause lensed images \citep{1937ApJ....86..217Z}, and it even
been argued that the phenomenon could help measure cosmological
parameters \citep{1964MNRAS.128..307R,1966MNRAS.132..101R}.  But
apparently nobody was expecting that lenses would need detailed
modelling.  In hindsight, the reason is not difficult to understand.
The scale of image separations is set by the angular Einstein radius
\begin{equation}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2}
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  For
stellar-mass lenses, the image separation is much larger than the size
of the lensing mass, so lenses are well approximated by point masses.
In galaxy lenses, the image separation is comparable to the size of
the galaxy; typically the lensed images are seen through the galaxy
halo.  Hence the lensed images depend on the detailed mass
distribution of the lensing galaxy.  Once observations of the lensed
images became available, it was obvious that mass modelling was
needed.

Since that early days, more than 400 secure lenses have been found
(see Figure \ref{fig:masterlens}) and detailed modelling of the mass
distribution is part of any research using multiply-imaging lenses.
But so far no modelling study has used anything approaching 400
lenses.  The largest single one \citep{2009ApJ...703L..51K} used 58
lenses to uncover the distribution of dark matter around galaxies.
\cite{2011ApJ...740...97L} combined lens models of 21 galaxies with
models of their stellar populations, to infer the relation between
stars and dark matter in galaxies.  \cite{2014MNRAS.437..600S}
modelled 18 time-delay lenses together to infer cosmological
parameters.

\begin{figure}
\centering
\includegraphics[width=\hsize]{fig/lenssky}
\caption{Sky distribution of 423 published lenses considered secure
  (from the Masterlens catalog at the University of Utah, maintained
  by Joel Brownstein and Leonidas Moustakas.) The map is in
  Hammer-Aitoff projection, with North up and $\rm RA=0$ in the
  middle. The empty swathes to the left and right of center are the
  Milky Way. {\em Should we drop this figure?  It's not relevant to
    modelling, but as there seems to be no publicly available summary
    of all known lenses, it does help set context.}}
\label{fig:masterlens}
\end{figure}

Integrating modelling into the lens discovery process would be a great
advantage.  In this paper, we develop a technique towards this goal,
working with the \sw lens search (Marshall et al, in prep, More et al,
in prep).  \sw is a citizen science project\footnote{\tt
  http://www.spacewarps.org} in which volunteers are presented
sky-survey images and invited to identify lens candidates.  Simulated
lenses are mixed in with the data, both to help train volunteers on
what to look for, and to estimate completeness and purity.  The
motivation for \sw is to enable volunteers, some of whom had
previously serendipitously identified lens candidates on earlier
citizen-science surveys, to make discoveries missed in automatic
searches by software robots.  Robots are very good with clean lensing
system in uncrowded fields with high signal-to-noise, but in general
test situations \citep{2009ApJ...694..924M} robots miss lenses (low
completeness) or contaminate the results with non-lenses (low purity).

The first tranche of survey data, introduced from May 2013, is an area
of $\sim172$ square degrees (or a quarter of a percent of the sky)
from the Canada-France-Hawaii Telescope Legacy Survey.  The survey is
divided into patches of $440\times 440$ pixels, each of which is seen
by ten volunteers.  These data had previously been analyzed by robots
\cite{Gavazzi2012, More2012ApJ}.
